Suitability of Graph Neural Network for Shape Memory Alloys

### 2024-05-24 src/tex/ms.tex
Fix script reference typo
Update to README and manuscript.

%% Packages
\usepackage{orcidlink} \usepackage[utf8]{inputenc}
\usepackage{fontspec} \usepackage{fontawesome5} \usepackage{amssymb}
\usepackage{amsmath} \usepackage{siunitx}
\usepackage[version=4]{mhchem} \usepackage{pdflscape}
%% Handles multipage table. Not sure allowed by pub.
\usepackage{longtable} \usepackage{tabularx} \usepackage{subcaption}
\usepackage{listings} \lstset{ basicstyle=\ttfamily,
  columns=fullflexible, frame=single, breaklines=true,
  postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space}, }

@@

% Abstract with filler text
\begin{abstract}
	Shape Memory Alloys (SMAs) such as NiTi are pivotal in vibrational
	dampening,robotic arms, prosthetic hands, and mechanical vales as a
	result of their unique displacive phase transformation
	properties. However, assessing their phonon characteristics, which
	provide insight into the displacive phase transformation, by using
	conventional Density Functional Theory (DFT) methods is
	computationally challenging. In this paper we assess the use of
	recent Graph Neural Network (GNN) potentials as a viable
	alternative. These GNN potentials have leveraged large DFT-computed
	databases for training and provide a good approximation to the
	potential energy surfaces which enable inference on energies and
	forces in lattice and atomic dynamic calculations. We compare GNN
	potentials with the Embedded Atom Method (EAM) and Modified EAM
	style potentials in predicting phonon properties of NiTi and
	PtTi. We find that in general the EAM/MEAM potentials are slightly
	more performant than the GNN in that they appear to more accurately
	describe equation of state, phonon dispersions, and elastic
	constants. However, the MACE potential shows significant promise as
	an exploratory potential for designing and characterizing more
	complex SMAs. Our findings suggest GNN potentials as a promising
	initial avenue for advancing SMA research, in particular ternary and
	quaternary systems, upon fine-tuning of pre-trained GNN models with
	more relevant DFT data (i.e., force-constants) to permit better
	outcomes.
\end{abstract}

@@

The design and development of shape memory alloys (SMA) via
\textit{in-silico} methods would provide great value in advancing
structural, medical, and other technologies. In order to enable such
activity would require sampling from very large structural and
compositional spaces and evaluating material properties indicative of
displacive phase transformation phenomena. The use of density
functional theory calculations to obtain properties such as equations
of state, phonon disperisons, and elastic constants, can be done,
however, such calculations are computationally demanding and are
potentially prohibitive for smaller groups trying to develop new
SMAs. To alleviate such restriction, one can resort to classical
interatomic potentials which can greatly improve computational cost in
evaluating the energies and forces more. However, the immediate
challenge is to whether a suitable interatomic potential exist given a
chemical system, and moreover, how transferable is it to different
compositions and phases. \par

Recently, there as been a growing effort in the development of machine
learning potentials. These deviate from traditional physics approaches
to interatomic potentials, which have historically been inspired from
more accurate methods like tight-binding method. In machine learning
potentials, one describes the atoms and structure using descriptors of
features, these are typically learned or prescribed. The features can
be prescribed in isolation or combination of structural,
compositional, and property characteristics of the material data
point. Futhermore, one may choose to fined new representations via
unsupervised learning. \par

The subsequent step once the featurization process is completed is to
determine a suitable neural network (NN) architecture which can learn
the target property or properties given the feature inputs. In the
case of atomistic simulations the target is prediction of the
potential energy surface and corresponding gradients (i.e.,
forces). Two of the most early attempts to do this were the works of
\citet{Bartok2010} and \citet{Thompson2015}, whereby projection onto
higher-dimensional basis, such as bispectrum components of local
neighbor density, is learned in order to describe the atomic
environments. The energy per atom is then determined by these learned
features by employing linear regression or non-linear Gaussian process
regression. We note that these are just two of the early examples but
similar potentials have been developed in earlier, in parallel, or
recently by various groups
\cite{Behler2007,Wang2018,Deringer2019,Lee2019}. \par

While these early attempts using ML models were successfully, they may
only include limited representations of structural and/or chemical
details to learn new features from which could limit
transferability. Futhermore, the training process was typically
restricted to a limited number of elemental constituents of the target
material system. As a consequence, for each chemical system and phases
of interest one needs to provide a suitable training dataset. To move
beyond such limitations, groups began using graph neural network (GNN)
architectures and message passing scheme to represent the atomic
structures as nodes and edges \cite{Reiser2022}. This enabled two
distinctions, first being that the nodes and edges corresponding to
the atoms and bonds could include features, and secondly the learning
could be done over a large structural and compositional space enabling
a generalization\footnote{Typically these are referred to as Universal
	interatomic potentials or forcefields since they in theory
	generalize well to any chemical system or phase.}. The use of GNNs
is efficient in representing structure-composition-property inputs to
additional NN layers for predicting potential energy surfaces and
gradients and was initially pursued in the CGCNN, MEGNet, and ALIGNN
potentials \cite{Choudhary2021,Chen2019,Xie2018}. The differences
between the approaches are predominately in atomic and bond features,
nearest-neighbor graph, and specific NN architecture. \par

The training of these early GNN utilized large density functional
theory calculation repositories and mostly tuned the learning cost
function to ground-state structure and energy. This resulted in
mean-absolute-errors and other measures of performance that could be
improved upon \cite{Riebesell2023}. More recently there have been
significant improvements in both training data used and the GNN
architecture that show promise for truly approaching a general
interatomic potential suitable for static and dynamic atomistic
calculations of multi-components and multi-phases. If successful,
identifying and designing new potential SMA materials using GNN
interatomic potentials could have tremendous impact on
technology. Therefore, the primary goal is to provide a qualitative
assessment of pre-trained GNN interatomic potentials that are promoted
as being transferable across compositions and phases. We specifically
focus on the predictions for equations of state, phonon bandstructure,
and elastic constants for various phases of NiTi and PtTi using three
GNN potentials: M3GNet, CHGNet, and MACE
\cite{Chen2022,Deng2023,Batatia2022}. We note that the
\citet{Riebesell2023} provides a summarized performance over these GNN
using the materials project database, but does not specifically
address SMA and phonon characteristics. \par

@@

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible,
interoperable and reusable (FAIR) principles\cite{Walsh2024}, we
utilize \showyourwork python workflow\cite{Luger2021} and several open
source python packages. Instructions for reproducing calculated data,
figures, tables, and manuscript are provided in
appendix~\ref{sec:appx_reproduce} and corresponding github
repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}. We
note that in order to achieve the different computing environments
required by the interatomic potentials we leverage the ability to
specify the python computing environments using
\href{https://github.com/showyourwork/showyourwork}{\showyourwork}
package.\par

\subsection{Calculation Details}
\label{sec:calc_details}
The ground-state structures for the NiTi phases \textit{B2},
\textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the
atomic simulation environment (ASE) package \cite{Larsen2017}. This
requires that the unit cell parameters, basis atoms, and spacegroup
number be provided. The initial unit cell parameters and basis
coordinates for these structures have been taken from the materials
project \cite{Jain2013} and several references
\cite{Haskins2016,Kadkhodaei2018}. The initial structures for NiTi are
shown in fig~\ref{fig:niti_structures} and are adapted for PtTi as
well given same space group symmetries. The energy and forces on the
atoms are determined using the ASE calculator interfaces for EAM,
MEAM, M3GNet, CHGNet, and
MACE.\cite{Mutter2010,Zhong2011,Ko2015,Kim2017,Chen2022,Deng2023,Batatia2022}
The EAM and MEAM potential interfaces are achieved through the use of
LAMMPS python library \cite{Thompson2022} and ASE. We have also
implemented calculators for potentials by \cite{Kavousi2019},
\cite{Wang2018}, and \cite{Choudhary2021} but are not shown in the
main manuscript of this work.\par

@@

\begin{figure}[ht!]
	\script{VisualizedStructures.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_VisualizedStructures.png}
		\caption{ Initial \ce{NiTi} structures (i.e., unrelaxed) used for
			(a) \textit{B2}(221), (b) \textit{B19}(11), (c)
			\textit{B19'}(51), and (d) \textit{BCO}(61) with spacegroup
			numbers are given in the parentheses. The green atoms are
			\ce{Ni} and \ce{Ti} as gray and unit cells are highlighted with
			parameters shown on axis. For all unit cells the $\alpha$ and
			$\gamma$ angles are 90.0$^\circ$. For PtTi, (a-b) structures are
			used with different unit cell parameters and basis coordinates
			and then optimized accordingly.  }
		\label{fig:niti_structures}
	\end{centering}
\end{figure}

@@

Shape memory alloys derive their unique displacive phase
transformations from the dynamic instability that occurs during the
application of temperature, strain, or stress. The origin of this
instability occurs due to softening or symmetry breaking of the phonon
modes in the phase. Therefore, improved understanding of the phonon
characteristics of the different phases that can occur for a given SMA
is particular important. One of the challenges is in performing such
calculations, which typically employ density functional theory (DFT)
in order to determine the atomic forces, which in turn are used to
construct the dynamical matrix. However, classical interatomic
potentials can alleviate the computational burden. \par

@@

In order to sample high-symmetry points in the irreducible Brillouin
zone (IBZ) for specific structures, we sample from locations specified
in the Bilbao crystallographic \textit{KVEC} utility\cite{Aroyo2014},
although the ASE structures object does provide a standardized routine
for sampling the irreducible Brillouin zone for the specific
structures.\cite{Setyawan2010}. We predominately focus on
high-symmetry points in the IBZ that correspond to the corners and
along directions known to exhibit phonon instabilities. The IBZ and
points symmetry point locations can be seen in appendix
\ref{sec:appx_ibz}. The final phonon dispersion diagrams are plot
along these points. \par

@@

Similar to the phonon calculations the elastic constants within the
harmonic approximation the elastic constants can be determined from
linear-response of strain-stress calculations. Therefore, the
calculation requires the use of crystal symmetry along with atomic
displacements to determine respective elastic constant components. We
use the \textit{Elastic}\cite{Jochym2018} python package which
provides routines for generating the proper atomic displacement matrix
given specific crystal space group. The elastic constants are
determined by solving a set of linear equations using singular value
decomposition and given calculated strains and stress. The
displacements occur with a min and max of 0.5\% and a total of 10
points are used. \par

@@

The EAM and MEAM potentials were developed to go beyond pair
interatomic potentials to treat point and extended defects in metal
and alloys systems. The motivation was taken from the physical results
of density functional theory and tight-binding methods, namely to add
to the pair potentials a energy functional which depends on the local
electron density which in turn is affected by the atomic
environment. These type of interatomic potentials are broadly referred
to as EAM or MEAM, with the later included direction (i.e. angle)
dependent interactions. \par

@@

\begin{figure}[ht!]
	\script{PlotCohesiveEnergy.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_CohesiveEnergyPlot.png}
		\caption{ Comparison of predicted \ce{NiTi} cohesive energy for
			respective ground-state structure and model. The experimental
			data point has been extracted from ref.~\cite{Vandermause2024}
			and the Materials Project entry for the \ce{NiTi} \textit{B2}
			structure \cite{MP--571} has also been included.  }
		\label{fig:ecoh_niti}
	\end{centering}
\end{figure}

For the \ce{PtTi} system we only consider the \textit{B2} and
\textit{B19'} structures. The MEAM potential shows a lower cohesive
energy than all three GNN. Interestingly the difference between the
MEAM and GNN is roughly the same order, $\approx 2.1$ eV, which maybe
appears to be related to the MP dataset. To our knowledge most the
pretrained GNN models are using the MP dataset which defaults to DFT
calculation results that employ the meta-GGA R2SCAN functional. This
could be the reason for the difference as most EAM and MEAM potentials
are either parameterized from experimental or DFT data using the PBE
functional\cite{Mutter2010,Ko2015}. \par

\begin{figure}[ht!]
	\script{PlotCohesiveEnergy.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/PtTi_CohesiveEnergyPlot.png}
		\caption{ Comparison of predicted \ce{PtTi} cohesive energy for
			respective ground-state structure and model.  }
		\label{fig:ecoh_ptti}
	\end{centering}
\end{figure}

@@

The MACE potential shown in fig.~\ref{fig:eos}(e) has a very different
set of EOS curves compard to the M3GNet and CHGNet. Here many of the
phases exhibit a stablity window, except the \textit{B33} phase, which
was stable in M3GNet and CHGNet. Interestingly, the \text{B19} phase,
not the \text{B19P} phase is shown to be the stable zero strain
phase. One unique perspective of the the UIP-GNN models is that they
provide a different displacive phase transformation pathway then that
shown by the EAM potentials. However, this does not include the
meta-stable R-phase or other potential meta-stable phases which can
also offer such transformation mechanisms. \par

Finally, the ALIGNN potential EOS was generated, but a few issues
arose during the calculations. The most prominant was the difficulty
in convergence of the structural optimzation to the specified
tolerance (see \ref{sec:methods}. Secondly the calculated cohesive
energies differs from the UIP-GNN by a factor $\approx 2$ and by $1$
eV for the EAM potentials. The \textit{B19P} phase is the stable at
zero strain and for a considerable portion of the compressive, but all
of the tensile. At compressive strain beyond $\approx 0.94$ the
\textit{B2} phase is stable.

\begin{figure}[ht!]
	\script{PlotEOS.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_EquationOfStates.png}
		\caption{ The equation of state curves for different \ce{NiTi}
			structures for (a) Mutter (EAM) (b) Zhong (EAM) (c) Ko
			(2NN-MEAM) (d) M3GNet (e) CHGNet (f) MACE interatomic
			potentials. The x-axis is the ratio of the isotropic strained
			unit cell volume, $V$, with respect to the ground-state unit
			cell volume, $V_o$.  }
		\label{fig:eos}
	\end{centering}
\end{figure}

@@

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B2_ModelsPhononBandstructures.png}
		\caption{ Ground-state, no isotropic strain, Phonon bandstructure
			of B2 phase for all models. The Mutter and Zhong EAM potentials
			differ in that Mutter predicts no dynamic instability of the B2
			phase at 0K.  }
		\label{fig:allmodels_b2}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B19_ModelsPhononBandstructures.png}
		\caption{ Ground-state phonon bandstructure of B19 phase for all
			models.  }
		\label{fig:allmodels_B19P}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B19P_ModelsPhononBandstructures.png}
		\caption{ Ground-state phonon bandstructure of
			\textit{B19}$^\prime$ phase for all models.  }
		\label{fig:allmodels_B19P}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_BCO_ModelsPhononBandstructures.png}
		\caption{ Ground-state phonon bandstructure of BCO phase for all
			models.  }
		\label{fig:allmodels_B19P}
	\end{centering}
\end{figure}

@@

The B2 phase of NiTi exhibits notable phonon instabilities,
particularly at the \textit{M} point in the Brillouin zone. These
instabilities are characterized by imaginary modes, represented as
negative phonon frequencies, indicating a dynamical instability in
this crystal structure. The \textit{M} point, located at the edge of
the Brillouin zone, corresponds to a wave vector of
$\left[\frac{1}{2}, \frac{1}{2}, 0\right]$ in units of the reciprocal
lattice vectors. The phonon dispersion curves for \ce{NiTi} in the B2
phase reveal a softening of the acoustic modes at the \textit{M}
point, suggesting a propensity for a structural phase transition. \par

@@

\section{Conclusion}
\label{sec:conclusion}
This word aimed to provide an assement of the of GNN univeral
interatomic potentials on the ability to capture the stable
ground-state phases of binary and ternary SMA. In addition, the
dynamic stability of these SMA is crucially evaluated to understand
the suitability in observing the onset of displacive phase
transformations via the softening of TA branch modes. In the case of
NiTi we find that all the GNN potentials predict either the Monoclinic
or BCO phases as stable over the B2 structure. However, not all agree
with the recent works by (cite) which have indicated that it is the
BCO phase which is the ground-state structure.

>>>>

%% Packages
\usepackage{orcidlink} \usepackage[utf8]{inputenc}
\usepackage{fontspec} \usepackage{fontawesome5} \usepackage{amssymb}
\usepackage{amsmath} \usepackage{siunitx}
\usepackage[version=4]{mhchem} \usepackage{pdflscape}
%% Handles multipage table. Not sure allowed by pub.
\usepackage{longtable} \usepackage{tabularx} \usepackage{subcaption}
\usepackage{listings} \lstset{basicstyle=\ttfamily,
  columns=fullflexible, frame=single, breaklines=true,
  postbreak=\mbox{\textcolor{red}{$\hookrightarrow$}\space},}

@@

% Abstract with filler text
\begin{abstract}
	Shape Memory Alloys (SMAs) such as NiTi are pivotal in vibrational
	dampening, robotic arms, prosthetic hands, and mechanical vales as a
	result of their unique displacive phase transformation
	properties. However, assessing their phonon characteristics, which
	provide insight into the displacive phase transformation, by using
	conventional Density Functional Theory (DFT) methods is
	computationally challenging. In this paper we assess the use of
	recent Graph Neural Network (GNN) potentials as a viable
	alternative. These GNN potentials have leveraged large DFT-computed
	databases for training and provide a good approximation to the
	potential energy surfaces which enable inference on energies and
	forces in lattice and atomic dynamic calculations. We compare GNN
	potentials with the Embedded Atom Method (EAM) and Modified EAM
	style potentials in predicting phonon properties of NiTi and
	PtTi. We find that in general the EAM/MEAM potentials are slightly
	more performant than the GNN in that they appear to more accurately
	describe equation of state, phonon dispersions, and elastic
	constants. However, the MACE potential shows significant promise as
	an exploratory potential for designing and characterizing more
	complex SMAs. Our findings suggest GNN potentials as a promising
	initial avenue for advancing SMA research, in particular ternary and
	quaternary systems, upon fine-tuning of pre-trained GNN models with
	more relevant DFT data (i.e., force-constants) to permit better
	outcomes.
\end{abstract}

@@

The design and development of shape memory alloys (SMA) via
\textit{in-silico} methods would provide great value in advancing
structural, medical, and other technologies. In order to enable such
activity would require sampling from very large structural and
compositional spaces and evaluating material properties indicative of
displacive phase transformation phenomena. The use of density
functional theory calculations to obtain properties such as equations
of state, phonon band structures, and elastic constants, can be done,
however, such calculations are computationally demanding and are
potentially prohibitive for smaller groups trying to develop new
SMAs. To alleviate such restriction, one can resort to classical
interatomic potentials which can greatly improve computational cost in
evaluating the energies and forces more. However, the immediate
challenge is to whether a suitable interatomic potential exist given a
chemical system, and moreover, how transferable is it to different
compositions and phases. \par

Recently, there as been a growing effort in the development of machine
learning potentials. These deviate from traditional based physics approaches
to interatomic potentials, which have historically been inspired from
more accurate methods like tight-binding method \cite{Finnis2010}. In machine learning
potentials, one describes the atoms and structure using descriptors (also called
features and are used interchangeably here on out) which are typically learned or prescribed.  If prescribed, the descriptors can
be isolated in terms of structure or composition, or a combination of the two. If needed properties can also be used as part of the descriptor set.\
compositional, and property characteristics of the material data
point. Futhermore, one may choose to fine new representations via
unsupervised learning \cite{Choudhary2021}. \par

The subsequent step, once the featurization process is completed, is to
determine a suitable machine learning model which can learn
the target property or properties given the descriptor inputs. For large data sets Neural Networks have become the preferred choice. \par

In the case of atomistic simulations the target is prediction of the
potential energy surface and corresponding gradients (i.e.,
forces). Two of the most early ML models to do this were the works of
\citet{Bartok2010} and \citet{Thompson2015}, whereby projection onto
higher-dimensional basis, such as bispectrum components of local
neighbor density, is learned in order to describe the atomic
environments. The energy per atom is then determined by these learned
features by employing linear regression or non-linear Gaussian process
regression. We note that these are just two of the early examples but
similar potentials have been developed in earlier, in parallel, or
recently by various groups
\cite{Behler2007,Wang2018,Deringer2019,Lee2019}. \par

While these early attempts using ML models were successfully, they may
only include limited representations of structural and/or chemical
details to learn new features from which could limit
transferability. Futhermore, the training process is typically
restricted to a limited number of elemental constituents of the target
material system. As a consequence, for each chemical system and phases
of interest one needs to provide a suitable dataset to learn from (i.e., training). To move beyond such limitations, groups began using graph neural network (GNN)
architectures and message passing scheme to represent the atomic
structures as nodes and edges \cite{Reiser2022}. This enabled two
distinctions, first being that the nodes and edges corresponding to
the atoms and bonds could include features, and secondly the learning
could be done over a large structural and compositional space enabling
a generalization\footnote{Typically these are referred to as Universal
	interatomic potentials or force fields since they in theory
	generalize well to any chemical system or phase.}. The use of GNNs
is efficient in representing structure-composition-property inputs to
additional NN layers for predicting potential energy surfaces and
gradients and was initially pursued in the CGCNN, MEGNet, and ALIGNN
potentials \cite{Choudhary2021,Chen2019,Xie2018}. The differences
between the approaches are predominately in atomic and bond features,
nearest-neighbor graph, and specific NN architecture. \par

The training of these early GNN utilized large density functional
theory calculation repositories and mostly tuned the learning cost
function to ground-state structure and energy. This resulted in
mean-absolute-errors and other measures of performance that could be
improved upon \cite{Riebesell2023}. More recently there have been
significant improvements in both training data used and the GNN
architecture that show promise for truly approaching a general
interatomic potential suitable for static and dynamic atomistic
calculations of multi-components and multi-phases. If successful,
identifying and designing new potential SMA materials using GNN
interatomic potentials could have tremendous impact on
technology. \par

The primary goal of this work is to provide a qualitative
assessment of pre-trained GNN interatomic potentials that are promoted
as being transferable across compositions and phases. We specifically
focus on the predictions for equations of state, phonon bandstructure,
and elastic constants for various phases of NiTi and PtTi using three
GNN potentials: M3GNet, CHGNet, and MACE
\cite{Chen2022,Deng2023,Batatia2022}. We note that the
\citet{Riebesell2023} provides a summarized performance for all of these GNN
using the materials project database\cite{Jain2013}, but does not specifically
address SMA and phonon characteristics. \par

@@

\subsection{Workflow}
\label{sec:workflow}
To promote reproducible research and align with findable, accessible,
interoperable and reusable (FAIR) principles\cite{Walsh2024}, we
utilize \showyourwork python workflow\cite{Luger2021} and several open
source python packages. Instructions for reproducing calculated data,
figures, tables, and manuscript are provided in
appendix~\ref{sec:appx_reproduce} and corresponding github
repo\footnote{\url{https://github.com/stefanbringuier/SMA_Phonons_GNNIP}}. We
note that in order to achieve the different computing environments
required by the interatomic potentials we leverage the ability to
specify virtual/containerized python computing environments using
\href{https://github.com/showyourwork/showyourwork}{\showyourwork}
package.\par

\subsection{Calculation Details}
\label{sec:calc_details}
The ground-state structures for the NiTi phases \textit{B2},
\textit{B19}, \textit{B19'}, and \textit{BCO} are generated using the
atomic simulation environment (ASE) package \cite{Larsen2017}. This
requires that the unit cell parameters, basis atoms, and spacegroup
number be provided. The initial unit cell parameters and basis
coordinates for these structures have been taken from the materials
project \cite{Jain2013} and several references
\cite{Haskins2016,Kadkhodaei2018}. The initial structures for NiTi are
shown in fig~\ref{fig:niti_structures} and are adapted for PtTi as
well, given same space group symmetries. The energy and forces on the
atoms are determined using the ASE calculator interfaces for EAM,
MEAM, M3GNet, CHGNet, and
MACE.\cite{Mutter2010,Zhong2011,Ko2015,Kim2017,Chen2022,Deng2023,Batatia2022}
The EAM and MEAM potential interfaces are achieved through the use of
LAMMPS python library \cite{Thompson2022} and ASE\@. We have also
implemented calculators for potentials by \cite{Kavousi2019},
\cite{Wang2018}, and \cite{Choudhary2021} but are not shown in the
main manuscript of this work.\par

@@

\begin{figure}[ht!]
	\script{VisualizeStructures.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_VisualizedStructures.png}
		\caption{Initial \ce{NiTi} structures (i.e., unrelaxed) used for
			(a) \textit{B2}(221), (b) \textit{B19}(11), (c)
			\textit{B19'}(51), and (d) \textit{BCO}(61) with spacegroup
			numbers are given in the parentheses. The green atoms are
			\ce{Ni} and \ce{Ti} as gray and unit cells are highlighted with
			parameters shown on axis. For all unit cells the $\alpha$ and
			$\gamma$ angles are 90.0$^\circ$. For PtTi, (a-b) structures are
			used with different unit cell parameters and basis coordinates
			and then optimized accordingly.  }
		\label{fig:niti_structures}
	\end{centering}
\end{figure}

@@

Shape memory alloys derive their unique displacive phase
transformations from the dynamic instability that occurs during the
application of temperature, strain, or stress. The origin of this
instability occurs due to softening or symmetry breaking of the phonon
modes in the phase. Therefore, improved understanding of the phonon
characteristics of the different phases that can occur for a given SMA
is particular important. \par

@@

In order to sample high-symmetry points in the irreducible Brillouin
zone (IBZ) for specific structures, we sample from locations specified
in the Bilbao crystallographic \textit{KVEC} utility\cite{Aroyo2014},
although the ASE structures object does provide a standardized routine
for sampling the irreducible Brillouin zone for the specific
structures \cite{Setyawan2010}. We predominately focus on
high-symmetry points in the IBZ that correspond to the corners and
along directions known to exhibit phonon instabilities. The IBZ and
points symmetry point locations can be seen in appendix~\ref{sec:appx_ibz}. The final phonon dispersion diagrams are plot
along these points. \par

@@

Similar to the phonon calculations the elastic constants within the
harmonic approximation, can be determined from
linear-response of strain-stress calculations. Therefore, the
calculation requires the use of crystal symmetry along with atomic
displacements to determine respective elastic constant components. We
use the \textit{Elastic}\cite{Jochym2018} python package which
provides routines for generating the proper atomic displacement matrix
given specific crystal space group. The elastic constants are
determined by solving a set of linear equations using singular value
decomposition and given calculated strains and stress. The
displacements occur with a min and max of 0.5\% and a total of 10
points are used. \par

@@

The EAM and MEAM potentials were developed to go beyond simpler pair
interatomic potentials to treat point and extended defects in metal
and alloys systems. The motivation was taken from the physical results
of density functional theory and tight-binding methods. The approach was to add
to the pair potentials a energy functional which depends on the local atomic
environment. These type of interatomic potentials are broadly referred
to as EAM or MEAM, with the later included direction
dependent interactions (i.e., angular terms). \par

@@

\begin{figure}[ht!]
	\script{PlotCohesiveEnergy.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_CohesiveEnergyPlot.png}
		\caption{Comparison of predicted \ce{NiTi} cohesive energy for
			respective ground-state structure and model. The experimental
			data point has been extracted from ref.~\cite{Vandermause2024}
			and the Materials Project entry for the \ce{NiTi} \textit{B2}
			structure \cite{MP--571} has also been included.  }
		\label{fig:ecoh_niti}
	\end{centering}
\end{figure}

For the \ce{PtTi} system we only consider the \textit{B2} and
\textit{B19'} structures. The MEAM potential shows a lower cohesive
energy than all three GNN\@. Interestingly the difference between the
MEAM and GNN is roughly the same order, $\approx 2.1$ eV, which maybe
appears to be related to the MP dataset. To our knowledge most the
pretrained GNN models are using the MP dataset which defaults to DFT
calculation results that employ the meta-GGA R2SCAN functional. This
could be the reason for the difference as most EAM and MEAM potentials
are either parameterized from experimental or DFT data using the PBE
functional\cite{Mutter2010,Ko2015}. \par

\begin{figure}[ht!]
	\script{PlotCohesiveEnergy.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/PtTi_CohesiveEnergyPlot.png}
		\caption{Comparison of predicted \ce{PtTi} cohesive energy for
			respective ground-state structure and model.  }
		\label{fig:ecoh_ptti}
	\end{centering}
\end{figure}

@@

The MACE potential shown in fig.~\ref{fig:eos}(e) has a very different
set of EOS curves compared to the M3GNet and CHGNet. Here many of the
phases exhibit a stability window, except the \textit{B33} phase, which
was stable in M3GNet and CHGNet. Interestingly, the \text{B19} phase,
not the \text{B19P} phase is shown to be the stable zero strain
phase. One unique perspective of the the UIP-GNN models is that they
provide a different displacive phase transformation pathway then that
shown by the EAM potentials. However, this does not include the
meta-stable R-phase or other potential meta-stable phases which can
also offer such transformation mechanisms. \par

Finally, the ALIGNN potential EOS was generated, but a few issues
arose during the calculations. The most prominent was the difficulty
in convergence of the structural optimzation to the specified
tolerance (see~\ref{sec:methods}). Secondly the calculated cohesive
energies differs from the UIP-GNN by a factor $\approx 2$ and by $1$
eV for the EAM potentials. The \textit{B19P} phase is the stable at
zero strain and for a considerable portion of the compressive, but all
of the tensile. At compressive strain beyond $\approx 0.94$ the
\textit{B2} phase is stable.

\begin{figure}[ht!]
	\script{PlotEOS.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_EquationOfStates.png}
		\caption{The equation of state curves for different \ce{NiTi}
			structures for (a) Mutter (EAM) (b) Zhong (EAM) (c) Ko
			(2NN-MEAM) (d) M3GNet (e) CHGNet (f) MACE interatomic
			potentials. The x-axis is the ratio of the isotropic strained
			unit cell volume, $V$, with respect to the ground-state unit
			cell volume, $V_o$.  }
		\label{fig:eos}
	\end{centering}
\end{figure}

@@

In the context of NiTi alloys, phonon instabilities play a crucial role in understanding the phase stability and transformations. The  phonon band structures for \textit{B2} phase, as shown in Fig.~\ref{fig:allmodelsB2}, reveal that the Mutter EAM potential predicts no dynamic instability at 0K. In contrast the Zhong EAM potential predicts onset of dynamic instability. In contrast, the Ko and MACE models exhibit phonon softening along the $\Gamma-M$ direction, aligning well with DFT results \cite{Wu2022}. The M3GNet and CHGNet models predict also predict mode instability but qualitatively differs significantly. This suggests that Ko and MACE models more accurately capture the intrinsic instability of the \textit{B2} phase, crucial for predicting the martensitic transformation to \textit{B19}$\prime$ or \textit{BCO} phases.

For the \textit{B19} phase, Fig.~\ref{fig:allmodelsB19}, the phonon band structures indicate instability in the $\Gamma-Y/B$ directions, consistent with DFT predictions. The Ko, CHGNet, and MACE models display similar trends, albeit with slight discrepancies. Furthermore, the \textit{B19}$\prime$ and \textit{BCO} phases, illustrated in Fig.~\ref{fig:allmodels_B19P} and Fig.~\ref{fig:allmodelsBCO}, respectively, highlight the stability predicted by DFT\@. Notably, only the Mutter, Zhong, and Ko models predict no mode instability for \textit{B19}$\prime$ and \textit{BCO}, while the others indicate potential mode instability. This reinforces the necessity here for fine-tuned GNN  potentials that can accurately capture the complex phonon dynamics and phase stability in NiTi alloys.

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B2_ModelsPhononBandstructures.png}
		\caption{Ground-state, no isotropic strain, Phonon bandstructure
			of B2 phase for all models. The Mutter (a) and Zhong (b) EAM potentials
			differ in that Mutter predicts no dynamic instability of the B2
			phase at 0K. The results for Ko (c), M3GNet (d), CHGNet (e), and MACE (f) all display the instability from $\Gamma-M$. The Ko (c) and MACE (f) models appear to agree best with DFT results \cite{Wu2022}.}
		\label{fig:allmodelsB2}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B19_ModelsPhononBandstructures.png}
		\caption{Ground-state phonon bandstructure of B19 phase for all
			models. Plots (a-f) follow same ordering as in fig.~\ref{fig:allmodelsB2}.The DFT results in ref.~\cite{Wu2022} indicate instability in $\Gamma-Y/B$ directions. In (c,e-f) the results are similar to the DFT but slight differ.}
		\label{fig:allmodelsB19}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_B19P_ModelsPhononBandstructures.png}
		\caption{Ground-state phonon bandstructure of
			\textit{B19}$^\prime$ phase for all models. Plots (a-f) follow same ordering as in fig.~\ref{fig:allmodelsB2}. As indicated by DFT this should be a potential stable structure however only (a-c) show no mode instability.}
		\label{fig:allmodelsB19P}
	\end{centering}
\end{figure}

\begin{figure}[!htp]
	\script{PlotPhonons.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/NiTi_BCO_ModelsPhononBandstructures.png}
		\caption{round-state phonon bandstructure of BCO phase for all
			models. Plots (a-f) follow same ordering as in fig.~\ref{fig:allmodelsB2}. As indicated by DFT this should be a potential stable structure however only (a-c) show no mode instability.}
		\label{fig:allmodelsBCO}
	\end{centering}
\end{figure}

@@

The \textit{B2} phase of NiTi exhibits notable phonon instabilities,
particularly at the \textit{M} point in the Brillouin zone. These
instabilities are characterized by imaginary modes, represented as
negative phonon frequencies, indicating a dynamical instability in
this crystal structure. The \textit{M} point, located at the edge of
the Brillouin zone, corresponds to a wave vector of
$\left[\frac{1}{2}, \frac{1}{2}, 0\right]$ in units of the reciprocal
lattice vectors. The phonon dispersion curves for \ce{NiTi} in the B2
phase reveal a softening of the acoustic modes at the \textit{M}
point, suggesting a propensity for a structural phase transition. \par

@@

\section{Conclusion}
\label{sec:conclusion}
This work provides a first step in the assessment of pre-trained universal GNN
interatomic potentials on ability to model SMA materials. In addition, the
dynamic stability of these SMA is crucially evaluated through phonon band structures to understand the suitability in observing the onset of displacive phase
transformations via the softening modes. In the case of
NiTi we find that all the GNN potentials predict either the Monoclinic
or BCO phases as stable over the B2 structure. However, not all agree
with relatively recent works \cite{Haskins2016,Wu2022} that indicate \textit{BCO} or \textit{B33} phase is the ground-state structure.
